---
title: Do it With Quasi-Quotation
author: Jameel Alsalam
date: '2017-11-20'
slug: quasi-quotation-applications
draft: true
categories:
  - quasiquotation
tags:
  - rlang
---

_This post is one of a series of posts that I am doing about expressions, quasi-quotation, and tidy evaluation. This post shows some useful things you can actually do with these tools without attempting to explain what's going on. The companion post tries to establish a mental model for quasi-quotation by analogy to a meta-recipe for cake._

_The R language has a powerful set of features for manipulating expressions which has recently been extended by the tidyverse team in the rlang package. I am really excited about the power and generality of quasi-quotation, especially in the realms of higher-order functions and creating unique function APIs. That said, I would recommend R users feel very comfortable with functions and functional programming (e.g., `apply` and/or `map` families of functions) before attempting quasi-quotation._

So you've already made an initial attempt to learn quasi-quotation/tidy evaluation, and the hill to understanding these tools still looks high. You are asking "why bother?", and tempted to just stick with what you know until forced, kicking-and-screaming to do it the "new way." In the following I will try to provide some quick wins and also show a taste of the power of quasi-quotation.

```{r setup}
suppressPackageStartupMessages(library("tidyverse"))
library("rlang")
```


## Recode LOTS of values, with Quasi-Quotation

Say you are doing some data cleaning and you decide to try out [`dplyr::recode`](http://dplyr.tidyverse.org/reference/recode.html). The examples in the documentation give a plausible way to change a couple of values:

```{r}
# Recode values with named arguments
x <- sample(c("a", "b", "c"), 10, replace = TRUE)
dplyr::recode(x, a = "Apple")
```

But wait -- do you really have to manually type in EVERY value that you want to replace?!?! Nope. You need to create a named list, and then use the `!!!` operator (pronounced 'bang bang bang' or unquote-splice).

```{r}
fruits <- list("Apple", "Banana", "Carrot")
fruit_lookup <- set_names(fruits, c("a", "b", "c"))

recode(x, !!!fruit_lookup)
```






