---
title: 'A Look at R Algo Optimization: Sieve of Eratosthenes'
author: Jameel
date: '2017-05-20'
slug: a-look-at-r-algo-optimization-sieve-of-eratosthenes
categories:
  - R
tags:
  - R
  - EulerProbs
---

In the course of working on problem #3 at [Project Euler](https://projecteuler.net/problem=3), I decided to take a detour to do a comparison of various implementations of the sieve of Eratosthenes. The problem asks:

> What is the largest prime factor of the number 600851475143 ?

I decided to begin approaching this by generating all primes less than this. So how do you generate the primes? The [Sieve of Eratosthenes](https://en.wikipedia.org/wiki/Sieve_of_Eratosthenes) jumps to mind:

> To find all the prime numbers less than or equal to a given integer n by Eratosthenes' method:
> 
> 1) Create a list of consecutive integers from 2 through n: (2, 3, 4, ..., n).
> 2) Initially, let p equal 2, the smallest prime number.
> 3) Enumerate the multiples of p by counting to n from 2p in increments of p, and mark them in the list (these will be 2p, 3p, 4p, ...; the p itself should not be marked).
> 4) Find the first number greater than p in the list that is not marked. If there was no such number, stop. Otherwise, let p now equal this new number (which is the next prime), and repeat from step 3.
> 5) When the algorithm terminates, the numbers remaining not marked in the list are all the primes below n.

Now, 6e11 is a pretty big number but... To put this in perspective, my computer has 16 GB of memory, which is only 16e9 bytes... but *whatever* (now that I'm this far, its obvious this is the long way around, but I'm having _fun_!). I have to admit that at this point I was feeling a bit intimidated, and decided to [google "sieve of eratosthenes R"](https://www.google.com/search?q=sieve+of+eratosthenes+r&oq=sieve+o&aqs=chrome.2.69i59j69i57j35i39j0l3.2233j0j7&sourceid=chrome&ie=UTF-8) to get started on the right foot.

[First hit](https://gist.github.com/seankross/5946396):

```{r sieve_first, echo=TRUE}
sieve_1 <- function(n){
  values <- rep(TRUE, n)
	values[1] <- FALSE
	last.prime <- 2
	for(i in last.prime:sqrt(n)){
		values[seq.int(2 * last.prime, n, last.prime)] <- FALSE
		last.prime <- last.prime + min(which(values[(last.prime + 1) : n]))
	}
	return(which(values))
}
```

And a more refined one within this [second hit](http://stackoverflow.com/questions/3789968/generate-a-list-of-primes-in-r-up-to-a-certain-number):

```{r sieve_second, echo=TRUE}
sieve_2 <- function(n)
{
  n <- as.integer(n)
  if(n > 1e9) stop("n too large")
  values <- rep(TRUE, n)
  values[1] <- FALSE
  last.prime <- 2L
  fsqr <- floor(sqrt(n))
  while (last.prime <= fsqr)
  {
    values[seq.int(2L*last.prime, n, last.prime)] <- FALSE
    sel <- which(values[(last.prime+1):(fsqr+1)])
    if(any(sel)){
      last.prime <- last.prime + min(sel)
    }else last.prime <- fsqr+1
  }
  which(values)
}

```

(note that I've modified the code in both to use the same names to make the remaining differences clear)

A quick performance comparison:

```{r microbenchmark, echo=TRUE}
library("microbenchmark")

microbenchmark(
  sieve_1(1e4),
  sieve_2(1e4),
  times = 10
)

```

Suffice it to say, the second one is way better. sieve_2 is running about 40X faster. I tried other values, and the differences are magnified the higher you get. Now, the two implementations are actually *very similar*, so what are the differences that are creating these large performance differences? Things I see:

1) sieve_2 uses a *while loop instead of a for loop*. This is important because you don't actually know how many iterations are going to be necessary before you start. sieve_1 includes a line that looks like maybe it is intended to jump forward when a prime is identified to the next non-prime, but it actually doesn't because once the for loop starts, you can't modify its counter.

A modified version of sieve_1 to instead use a while loop might look like this:

```{r sieve_1_while, echo=TRUE}
sieve_1_v2 <- function(n){
  values <- rep(TRUE, n)
	values[1] <- FALSE
	last.prime <- 2
	
	while(last.prime < sqrt(n)){
		values[seq.int(2 * last.prime, n, last.prime)] <- FALSE
		last.prime <- last.prime + min(which(values[(last.prime + 1) : n]))
	}
	return(which(values))
}

microbenchmark(
  sieve_1(1e4),
  sieve_1_v2(1e4),
  times = 10
)

```

This minor change has created an improvement of a factor of 3. To explain this, I tracked how many times each one iterates.

```{r iter_track, echo=FALSE}
sieve_1_tracked <- function(n){
  values <- rep(TRUE, n)
	values[1] <- FALSE
	last.prime <- 2
	iter <- 0
	
	for(i in last.prime:sqrt(n)){
		values[seq.int(2 * last.prime, n, last.prime)] <- FALSE
		last.prime <- last.prime + min(which(values[(last.prime + 1) : n]))
		iter <- iter + 1
	}
	
	print(paste0("sieve_1 iterated ", iter, " times."))
	return(which(values))
}

sieve_1_v2_tracked <- function(n){
  values <- rep(TRUE, n)
	values[1] <- FALSE
	last.prime <- 2
	iter <- 0
	
	
	while(last.prime < sqrt(n)){
		values[seq.int(2 * last.prime, n, last.prime)] <- FALSE
		last.prime <- last.prime + min(which(values[(last.prime + 1) : n]))
		iter <- iter + 1
	}
	
	print(paste0("sieve_1_v2 iterated ", iter, " times."))
	return(which(values))
}

sieve_2_tracked <- function(n)
{
  n <- as.integer(n)
  if(n > 1e9) stop("n too large")
  values <- rep(TRUE, n)
  values[1] <- FALSE
  last.prime <- 2L
  fsqr <- floor(sqrt(n))
  iter <- 0
	
  while (last.prime <= fsqr)
  {
    values[seq.int(2L*last.prime, n, last.prime)] <- FALSE
    sel <- which(values[(last.prime+1):(fsqr+1)])
    if(any(sel)){
      last.prime <- last.prime + min(sel)
      iter <- iter + 1
    }else last.prime <- fsqr+1
  }
  
  print(paste0("sieve_2 iterated ", iter, " times."))
  which(values)
}

x1 <- sieve_1_tracked(1e4)
x2 <- sieve_1_v2_tracked(1e4)
x3 <- sieve_2_tracked(1e4)  

```

I'm assuming there's a shmancy way to do this, but I just inserted counters inside the loops and then printed at the end. It was a little annoying, because I didn't want to mess up the functions, so I made alternate "tracked" versions of them. From here, what else can we improve?

2) Between the improved version and sieve_2, the better one pre-stores some values that are re-used. For example, sieve_2 only calculates sqrt(n) a single time prior to the loop beginning, whereas sieve_1_v2 does it for each iteration of the loop.

3) sieve_2 also specifically labels some of the variables as integers that might otherwise be doubles. This is a little awkward, because one strength of R is that you don't need to worry about types, but I guess once you start optimizing things, you actually do.

4) The formulas in sieve_2 seem a little bit different, like some pieces have been factored out to avoid repetition. This will take some closer looking. But I don't have the time for these right now.

Now, none of these are going to be good enough for Euler #3. To get all the primes up to where we need, we either need to split up the task into bits, or a different approach that works better for big numbers, or a different strategy entirely.  But all that is probaby for another time. I would like to say that I will update this post with some enhancements later on, but I don't fully know if blogdown lets you do that.

There are also some optimizations of the sieve algorithm. For example, you can start by omitting the even numbers, and I think something to do with multiples of 30. I don't immediately know how that would interect with the strategy these implementations are using of using a vector of logicals. More things to explore!

P.S., I wanted to highlight the specific changes that I made within the code, but I don't know how to do that within rmarkdown. There must be a way. Let me know if you know what it is! Or if you know a good way to go about updating posts. Like, do you need to change both the date field and the filename? Is there a way to use commit references from git? I dunno, but I can tell there are things to learn there.