---
title: "Quasi-Quotation w/Environments -> Time for Tidy-Eval"
author: "Jameel Alsalam"
date: '2017-11-20'
draft: yes
slug: multi-environments-tidy-eval
tags: []
categories:
- quasiquotation
- R
---

```{r setup, cache = F}
knitr::opts_chunk$set(error = TRUE)
```

## NSE functions are your Friend that wants the Recipe

So what can you do with the recipe other than bake it? You can pass it around. R has lots of special functions that are expecting to work on a combination of regular objects and expressions. For example, let's imagine you have a friend that has the ingredients for the frosting in their pantry. So you give them the cake you baked, along with the recipe for the frosting, and they give you back a fully frosted cake.

```{r}

# friend <- function(cake, frosting_recipe) {
#   #pantry
#   sugar <- 87
#   vanilla <- 10
#   butter <- 60
#   milk  <- 30
#   
#   frosting <- bake(frosting_recipe)
#   frosted_cake <- cake + frosting
# }
# 
# my_birthday_cake <- friend(cake, frosting_recipe)
# 
# objects()

```

Note that now we have a birthday cake, but we still don't have any milk or vanilla in our environment, even though those were necessary to evaluate the `frosting_recipe` expression. That's because those are in the environment of the `friend` function, not the global environment. (Remember: in this extended metaphor, the Environment is the Pantry. Your friend's pantry is not the same as your pantry.) (Apologies, I know I am pushing this comparison hard.)

Dplyr functions are functions that work on a combination of regular objects and expressions. The `data` argument is a regular object, but most of the others are expressions, not regular objects. You don't feel like you are passing in an expression because these functions use tricks to capture the expression you type even though you didn't explicitly write it down as a recipe. But I think this is a useful way to think about it.


## `quote()` v. `quo()` v. `expr()`: Which Pantry to Look In

To play out this example, we are going to start over, and distribute the task of making this cake to several friends. Friend1 has the necessary ingredients for the cake, Friend2 has the ingredients for the frosting, and the global environment has the candle.

```{r}
#starting over, so please clear the environment
rm(list = ls())
library(rlang)

friend1 <- function(){
  flour <- 20
  eggs  <- 50
  sugar <- 72
  
  cake_recipe_quo   <-   quo(flour + eggs + sugar)
  cake_recipe_expr  <-  expr(flour + eggs + sugar)
  cake_recipe_quote <- quote(flour + eggs + sugar)

  return(list(quo   = cake_recipe_quo, 
              expr  = cake_recipe_expr,
              quote = cake_recipe_quote))
}

friend2 <- function(){
  sugar <- 87
  vanilla <- 10
  butter <- 60
  milk  <- 30
  
  frosting_recipe_quo   <-   quo(sugar + vanilla + butter + milk)
  frosting_recipe_expr  <-  expr(sugar + vanilla + butter + milk)
  frosting_recipe_quote <- quote(sugar + vanilla + butter + milk)
  
  return(list(quo   = frosting_recipe_quo, 
              expr  = frosting_recipe_expr,
              quote = frosting_recipe_quote))
}

candle <- 0
cake_recipes     <- friend1()
frosting_recipes <- friend2()
```

The three versions **look** similar, but they have different scoping characteristics when we try to bake them.

```{r}
bake <- eval_tidy

bake(cake_recipes$quo)
```

The reason `quo` works here is that the quosure is not *just* the expression -- it also carries along information about the environment in which it was created ("lexical scoping"). The version `cake_recipes$quo` was created inside the friend1 function, where the ingredients are also defined. So it works fine.

```{r}
bake(cake_recipes$expr)
```

This variation gives an error because `expr` just creates a bare expression without any information about where to look for the ingredients. So when it is evaluated at the top level, R looks for `flour` in the global environment and doesn't find it.

```{r}
bake(cake_recipes$quote)

```

Similar problem here -- R doesn't look in the right place for the ingredients.

To finish everything off, let's show that we can not only rewrite a recipe for the full cake, but also pass the recipe to a third friend which doesn't have ANY ingredients in its local environment, but nonetheless that friend can evaluate the quosure.

```{r}
birthday_cake_recipe <- quo((!!cake_recipes$quo) + (!!frosting_recipes$quo) + candle)
print("The recipe is:")
birthday_cake_recipe

friend3 <- function(a_quosure){
  birthday_cake <- eval_tidy(a_quosure)
}

my_birthday_cake <- friend3(birthday_cake_recipe)
print("This is the actual cake, in terms of calories:")
my_birthday_cake
```
